Distribution-oblivious Programming with Dish

### 2019-11-20 pldi/draft.tex
Massive rewrite of background section

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}

@@

\subsection{Unix Pipelines, Informally}
\label{bg:pipelines}

A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
Behind the scenes, \unix facilitates program scheduling and communication synchronization.
Intermediate programs run concurrently consuming input and producing output;
  they can be developed in any programming language, as long as they conform to a common interface.
This interface allows any combination of the following:
  accepting input from the standard input stream (stdin),
  generating output to the standard output stream (stdout), and
  generating errors or diagnostics to the standard error stream (stderr)
These standard streams are lines of characters, with a few special characters---for example, EOF is a special end-of-file character signaling the end of a stream.
Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

The operation of commands can be tuned by two mechanisms, namely command-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  their parsing and evaluation is left entirely up to each command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
for example, given a file identifier, the stream redirection operator can direct stdout or stderr to the file ---of course, this can be a FIFO that is concurrently used as input by a different command.
More interestingly, some commands allow splitting a single stream into multiple streams while others allow merging streams together (either one after the other, or fusing (zipping) them column-wise into a single stream.

First, as described above, several commands can be redirected
Second,

\ttt{xargs}
\ttt{tee}

In combination, these constructs are powerful enough to introduce circular streams. % or stream cycles?

Explicit Redirection
Implicit splitting and merging 
This can be done through the file-system operators

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

@@

Special files are particularly interesting because they are resolved to OS structures.
One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
% Here are a few examples:
% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
Similarly, \unix exposes file-system handles for various ephmeral structures, including standard streams.

The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines---for example:

% These streams are created by the C library (glibc) at the start of program execution,
% and new streams can be created to connect to files, sockets, pipes,

This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.

Apart from pipes, the \unix shell features several  other forms of program composition.
Command substitution, enclosed in \ttt{\$()}, replaces the contents of the 
Code block redirects { - }, 
grouping command 
single redirection to redirect the standard output of both commands in the
this is analogous to an anonymous function, as it can be assigned to a variable and be manipulated as a subscript.
A subshell enclosed in \ttt{()} is similar , but executes the block in a new sub-shell---avoiding polluting environment. % but incurring higher overhead.

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat /proc/ >                            (($$p_4$$))
\end{lstlisting}

% For Unix commands, the newline character is special; as such, Unix streams can be seen as lazy lists of strings---lines of text separated by the newline character.

% Pipeline chaining goes hand-in-hand with the Unix toolbox philosophy: 
%   programs do one thing well, they work together, and they manipulate text streams as their universal interface.

@@

\heading{Basics}
The standard shell syntax for pipelines a list of commands separated by vertical bars (``pipes'', in Unix terminology).

that are not interpreted by the shell, but are local to each command.

Additional details augment this simple interface.
Commands can be specified with options arguments

\begin{figure}[t]
\centering
\includegraphics[width=0.45\textwidth]{./figs/dish_ex1.pdf}
\label{fig:example}
\end{figure}

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 ls -l | wc -l                           (($$p_2$$))
\end{lstlisting}

\heading{Generation}
Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat * | wc -l > out.txt                 (($$p_3$$))
\end{lstlisting}

\heading{Identifiers}

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat /proc/ >                            (($$p_5$$))
\end{lstlisting}

\heading{Composition}
anonymous pipes

Composition

The shell provides several ways of grouping commands and manipulating their
overall input and output channels

{ echo 1; echo 2; } > out.txt

>>>>

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{wrapfig}

@@

\subsection{Unix Composition, Informally}
\label{bg:pipelines}

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

Commands can be tuned by two mechanisms, namely command-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  their parsing and evaluation is left entirely up to each command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

@@

Special files are particularly interesting because they are resolved to OS structures.
One example is the special \ttt{/proc} filesystem (procfs) that presents information about processes and the operating system, offering a feature analogous to programmatic introspection:
  with \ttt{procfs} commands can inspect and alter their directory or the flags passed to them.
% Here are a few examples:
% (i) \ttt{/proc/PID/cmdline} contains the command that originally started the process,
% (ii) \ttt{/proc/modules} contains a list of the kernel modules currently loaded, and
% (iii) \ttt{/proc/PID/cwd} contains a symlink to the current working directory of the process.
Similarly, \unix exposes file-system handles for various ephemeral structures, including standard streams.
This bidirectional relationship between ephemeral streams and persistent names---the fact that streams can be named in the file-system and files can be converted to streams---is a key enabler of the runtime rebinding achieved during our transformations.

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
As described before, the file can be a FIFO dubbing as input to another command.
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

% Explicit Redirection, Implicit splitting and merging, This can be done through the file-system operators

The combination of file-system identifiers pointing to ephemeral streams and the stream redirection described above enable surprisingly expressive pipelines. %---for example:
The most interesting feature of these pipelines is the presence of cycles:
  a pipeline segment processes some input, part of which is extracting the next input to process.
One example use case, discussed in the evaluation section~\sx{eval} is a web crawler that feeds extracted URLs from the input stream back to the start of the pipeline.
% These streams are created by the C library (glibc) at the start of program execution,
% and new streams can be created to connect to files, sockets, pipes,

Apart from pipes, the \unix shell provides several other forms of program composition such as sequential (\ttt{;}) and parallel (\ttt{\&}) composition operators.
Command substitution, expressed as \ttt{\$($c$)}, replaces the expression with the results of executing $c$.
Code blocks, expressed as \ttt{\{$c_1; c_2$\}}, can redirect streams for the entire group of $c_1$ and $c_2$;
  this is analogous to an anonymous function, which contains many statements and can possibly be assigned to a variable that can later be invoked with parameters.
Subshells, enclosed in \ttt{($c_1; c_2$)}, are similar to code blocks but spawn a child shell, thus avoiding the pollution of their environment variable context. % but incurring higher overhead.

@@

% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat /proc/ >                            (($$p_4$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  ls -l | wc -l                           (($$p_2$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat * | wc -l > out.txt                 (($$p_3$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%  cat /proc/ >                            (($$p_5$$))
% \end{lstlisting}
% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
% { echo 1; echo 2; } > out.txt
% \end{lstlisting}

@@

% \heading{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

\subsection{Overview}
